Improving Outcome Predictions for Patients Receiving Mechanical Circulatory Support by Optimizing Imputation of Missing Values

### 2020-12-16 doc_arXiv/doc_arXiv.Rmd
word doc for circ

---
title: Improving Outcome Predictions for Patients Receiving Mechanical Circulatory Support by Optimizing Imputation of Missing Values
authors:
  - name: Byron C. Jaeger
    thanks: Source code available at https://github.com/bcjaeger/INTERMACS-missing-data
    department: Department of Biostatistics
    affiliation: University of Alabama at Birmingham
    location: Birmingham, AL
    email: bcjaeger@uab.edu
  - name: Ryan Cantor, PhD
    department: Kirklin Institute for Research in Surgical Outcomes
    affiliation: University of Alabama at Birmingham
    location: Birmingham, AL
  - name: Venkata Sthanam, MS
    department: Kirklin Institute for Research in Surgical Outcomes
    affiliation: University of Alabama at Birmingham
    location: Birmingham, AL
  - name: Rongbing Xie, DrPH
    department: Department of Surgery
    affiliation: University of Alabama at Birmingham
    location: Birmingham, AL
  - name: James K. Kirklin, MD
    department: Kirklin Institute for Research in Surgical Outcomes
    affiliation: University of Alabama at Birmingham
    location: Birmingham, AL
  - name: Ramaraju Rudraraju, PhD
    department: Kirklin Institute for Research in Surgical Outcomes
    affiliation: University of Alabama at Birmingham
    location: Birmingham, AL
    email: ramaraju@uab.edu
abstract: |
  \textbf{Background} Risk prediction models play an important role in clinical decision making. When developing risk prediction models, practitioners often impute missing values to the mean. We evaluated the impact of applying other strategies to impute missing values on the prognostic accuracy of downstream risk prediction models, i.e., models fitted to the imputed data. A secondary objective was to compare the accuracy of imputation methods based on artificially induced missing values. To complete these objectives, we used data from the Interagency Registry for Mechanically Assisted Circulatory Support (INTERMACS).  \newline\textbf{Methods and Results} We applied twelve imputation strategies in combination with two different modeling strategies for mortality and transplant risk prediction following surgery to receive mechanical circulatory support. Model performance was evaluated using Monte-Carlo cross validation and measured based on outcomes 6-months following surgery using the scaled Brier score, concordance index, and calibration error. We used Bayesian hierarchical models to compare model performance. Multiple imputation with random forests emerged as a robust strategy to impute missing values, increasing model concordance by 0.003 (25th, 75th percentile: 0.0008, 0.052) compared with imputation to the mean for mortality risk prediction using a downstream proportional hazards model. The posterior probability that single and multiple imputation using random forests would improve concordance versus mean imputation was 0.464 and >0.999, respectively. \newline\textbf{Conclusion} Selecting an optimal strategy to impute missing values and applying multiple imputation can improve the prognostic accuracy of downstream risk prediction models.
keywords:
  - missing data
  - INTERMACS
  - imputation
  - heart failure
  - mechanical circulatory support
  - risk prediction
bibliography: references.bib
output: rticles::arxiv_article
header-includes:
   - \newcommand{\ie}{\textit{i.e., }}
   - \newcommand{\eg}{\textit{e.g., }}
   - \newcommand{\cstat}{\widehat{\textrm{C}}(t)}
   - \newcommand{\bstat}{\widehat{\textrm{BS}}(t)}
   - \newcommand{\ibstat}{\mathcal{\widehat{BS}}(\tau)}
---

@@

The Interagency Registry  for  Mechanically  Assisted  Circulatory  Support (INTERMACS) was launched to improve MCS patient outcomes through the collection of patient characteristics, medical events, and long terms outcomes at a nation wide level. Currently, INTERMACS comprises over 20,000 patients who have received a MCS device. As the largest registry for data on patients receiving MCS devices, INTERMACS has been leveraged to develop numerous risk prediction models for mortality and other types of adverse events that may occur after receiving a device \cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society, Adamo950}. An important step for developing risk prediction models is dealing with missing data, which is common in a “real world” database that is not designed for the rigor of data  completeness found in a  clinical trial. While some missing values are related to data entry error, others may relate to instability of a patient’s circulatory status at the time of implant.  For instance, hemodynamic laboratory values are missing for patients in whom  invasive catheterizations were not performed. Patients without invasive catheterizations may be too ill (unstable) to tolerate the procedure. Instead, such patients may, after determination of basic hemodynamics, proceed directly to device placement.

@@

\paragraph{Conclusion} Selecting an optimal strategy to impute missing values can impact the prognostic accuracy of downstream risk prediction models. In the current analysis, conducting multiple imputation using random forests emerged as an optimal strategy to impute missing values in the INTERMACS data. This investigation can directly inform future analyses of INTERMACS data and provide evidence quantifying the benefit of imputing missing data with sound methodology. 
\paragraph{Sources of Funding} This work is supported by the American Heart Association, grant# 18AIML34280052. De-identified data provided from the INTERMACS contract with from the National Heart, Lung, and Blood Institute; National Institutes of Health; and Department of Health and Human Services, under contract# HHSN268201100025C.

>>>>

---
title: Improving Outcome Predictions for Patients Receiving Mechanical Circulatory Support by Optimizing Imputation of Missing Values
authors:
  - name: Byron C. Jaeger
    thanks: Source code available at https://github.com/bcjaeger/INTERMACS-missing-data
    department: Department of Biostatistics
    affiliation: University of Alabama at Birmingham
    location: Birmingham, AL
    email: bcjaeger@uab.edu
  - name: Ryan Cantor, PhD
    department: Kirklin Institute for Research in Surgical Outcomes
    affiliation: University of Alabama at Birmingham
    location: Birmingham, AL
  - name: Venkata Sthanam, MS
    department: Kirklin Institute for Research in Surgical Outcomes
    affiliation: University of Alabama at Birmingham
    location: Birmingham, AL
  - name: Rongbing Xie, DrPH
    department: Department of Surgery
    affiliation: University of Alabama at Birmingham
    location: Birmingham, AL
  - name: James K. Kirklin, MD
    department: Kirklin Institute for Research in Surgical Outcomes
    affiliation: University of Alabama at Birmingham
    location: Birmingham, AL
  - name: Ramaraju Rudraraju, PhD
    department: Kirklin Institute for Research in Surgical Outcomes
    affiliation: University of Alabama at Birmingham
    location: Birmingham, AL
    email: ramaraju@uab.edu
abstract: |
  \textbf{Background} Risk prediction models play an important role in clinical decision making. When developing risk prediction models, practitioners often impute missing values to the mean. We evaluated the impact of applying other strategies to impute missing values on the prognostic accuracy of downstream risk prediction models, i.e., models fitted to the imputed data. A secondary objective was to compare the accuracy of imputation methods based on artificially induced missing values. To complete these objectives, we used data from the Interagency Registry for Mechanically Assisted Circulatory Support (INTERMACS).  \newline\textbf{Methods and Results} We applied twelve imputation strategies in combination with two different modeling strategies for mortality and transplant risk prediction following surgery to receive mechanical circulatory support. Model performance was evaluated using Monte-Carlo cross validation and measured based on outcomes 6-months following surgery using the scaled Brier score, concordance index, and calibration error. We used Bayesian hierarchical models to compare model performance. Multiple imputation with random forests emerged as a robust strategy to impute missing values, increasing model concordance by 0.003 (25th, 75th percentile: 0.0008, 0.052) compared with imputation to the mean for mortality risk prediction using a downstream proportional hazards model. The posterior probability that single and multiple imputation using random forests would improve concordance versus mean imputation was 0.464 and >0.999, respectively. \newline\textbf{Conclusion} Selecting an optimal strategy to impute missing values such as random forests and applying multiple imputation can improve the prognostic accuracy of downstream risk prediction models.
keywords:
  - missing data
  - INTERMACS
  - imputation
  - heart failure
  - mechanical circulatory support
  - risk prediction
bibliography: references.bib
output: rticles::arxiv_article
header-includes:
   - \newcommand{\ie}{\textit{i.e., }}
   - \newcommand{\eg}{\textit{e.g., }}
   - \newcommand{\cstat}{\widehat{\textrm{C}}(t)}
   - \newcommand{\bstat}{\widehat{\textrm{BS}}(t)}
   - \newcommand{\ibstat}{\mathcal{\widehat{BS}}(\tau)}
---

@@

The Interagency Registry  for  Mechanically  Assisted  Circulatory  Support (INTERMACS) was launched to improve MCS patient outcomes through the collection of patient characteristics, medical events, and long terms outcomes at a nationwide level. Currently, INTERMACS comprises over 20,000 patients who have received a MCS device. As the largest registry for data on patients receiving MCS devices, INTERMACS has been leveraged to develop numerous risk prediction models for mortality and other types of adverse events that may occur after receiving a device \cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society, Adamo950}. An important step for developing risk prediction models is dealing with missing data, which is common in a “real world” database that is not designed for the rigor of data  completeness found in a  clinical trial. While some missing values are related to data entry error, others may relate to instability of a patient’s circulatory status at the time of implant.  For instance, hemodynamic laboratory values are missing for patients in whom  invasive catheterizations were not performed. Patients without invasive catheterizations may be too ill (unstable) to tolerate the procedure. Instead, such patients may, after determination of basic hemodynamics, proceed directly to device placement.

@@

\paragraph{Conclusion} Selecting an optimal strategy to impute missing values such as random forests can impact the prognostic accuracy of downstream risk prediction models. In the current analysis, conducting multiple imputation using random forests emerged as an optimal strategy to impute missing values in the INTERMACS data. This investigation can directly inform future analyses of INTERMACS data and provide evidence quantifying the benefit of imputing missing data with sound methodology. 
\paragraph{Sources of Funding} This work is supported by the American Heart Association, grant# 18AIML34280052. De-identified data provided from the INTERMACS contract with from the National Heart, Lung, and Blood Institute; National Institutes of Health; and Department of Health and Human Services, under contract# HHSN268201100025C.

### 2020-12-16 doc_arXiv/doc_arXiv.tex
word doc for circ

\begin{abstract}
\textbf{Background} Risk prediction models play an important role in
clinical decision making. When developing risk prediction models,
practitioners often impute missing values to the mean. We evaluated the
impact of applying other strategies to impute missing values on the
prognostic accuracy of downstream risk prediction models, i.e., models
fitted to the imputed data. A secondary objective was to compare the
accuracy of imputation methods based on artificially induced missing
values. To complete these objectives, we used data from the Interagency
Registry for Mechanically Assisted Circulatory Support (INTERMACS).
\newline\textbf{Methods and Results} We applied twelve imputation
strategies in combination with two different modeling strategies for
mortality and transplant risk prediction following surgery to receive
mechanical circulatory support. Model performance was evaluated using
Monte-Carlo cross validation and measured based on outcomes 6-months
following surgery using the scaled Brier score, concordance index, and
calibration error. We used Bayesian hierarchical models to compare model
performance. Multiple imputation with random forests emerged as a robust
strategy to impute missing values, increasing model concordance by 0.003
(25th, 75th percentile: 0.0008, 0.052) compared with imputation to the
mean for mortality risk prediction using a downstream proportional
hazards model. The posterior probability that single and multiple
imputation using random forests would improve concordance versus mean
imputation was 0.464 and \textgreater0.999, respectively.
\newline\textbf{Conclusion} Selecting an optimal strategy to impute
missing values and applying multiple imputation can improve the
prognostic accuracy of downstream risk prediction models.
\end{abstract}

@@

The Interagency Registry for Mechanically Assisted Circulatory Support
(INTERMACS) was launched to improve MCS patient outcomes through the
collection of patient characteristics, medical events, and long terms
outcomes at a nation wide level. Currently, INTERMACS comprises over
20,000 patients who have received a MCS device. As the largest registry
for data on patients receiving MCS devices, INTERMACS has been leveraged
to develop numerous risk prediction models for mortality and other types
of adverse events that may occur after receiving a device
\cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society, Adamo950}.
An important step for developing risk prediction models is dealing with
missing data, which is common in a ``real world'' database that is not
designed for the rigor of data completeness found in a clinical trial.
While some missing values are related to data entry error, others may
relate to instability of a patient's circulatory status at the time of
implant. For instance, hemodynamic laboratory values are missing for
patients in whom invasive catheterizations were not performed. Patients
without invasive catheterizations may be too ill (unstable) to tolerate
the procedure. Instead, such patients may, after determination of basic
hemodynamics, proceed directly to device placement.

@@

Selecting an optimal strategy to impute missing values can impact the
prognostic accuracy of downstream risk prediction models. In the current
analysis, conducting multiple imputation using random forests emerged as
an optimal strategy to impute missing values in the INTERMACS data. This
investigation can directly inform future analyses of INTERMACS data and
provide evidence quantifying the benefit of imputing missing data with
sound methodology.

>>>>

\begin{abstract}
\textbf{Background} Risk prediction models play an important role in
clinical decision making. When developing risk prediction models,
practitioners often impute missing values to the mean. We evaluated the
impact of applying other strategies to impute missing values on the
prognostic accuracy of downstream risk prediction models, i.e., models
fitted to the imputed data. A secondary objective was to compare the
accuracy of imputation methods based on artificially induced missing
values. To complete these objectives, we used data from the Interagency
Registry for Mechanically Assisted Circulatory Support (INTERMACS).
\newline\textbf{Methods and Results} We applied twelve imputation
strategies in combination with two different modeling strategies for
mortality and transplant risk prediction following surgery to receive
mechanical circulatory support. Model performance was evaluated using
Monte-Carlo cross validation and measured based on outcomes 6-months
following surgery using the scaled Brier score, concordance index, and
calibration error. We used Bayesian hierarchical models to compare model
performance. Multiple imputation with random forests emerged as a robust
strategy to impute missing values, increasing model concordance by 0.003
(25th, 75th percentile: 0.0008, 0.052) compared with imputation to the
mean for mortality risk prediction using a downstream proportional
hazards model. The posterior probability that single and multiple
imputation using random forests would improve concordance versus mean
imputation was 0.464 and \textgreater0.999, respectively.
\newline\textbf{Conclusion} Selecting an optimal strategy to impute
missing values such as random forests and applying multiple imputation
can improve the prognostic accuracy of downstream risk prediction
models.
\end{abstract}

@@

The Interagency Registry for Mechanically Assisted Circulatory Support
(INTERMACS) was launched to improve MCS patient outcomes through the
collection of patient characteristics, medical events, and long terms
outcomes at a nationwide level. Currently, INTERMACS comprises over
20,000 patients who have received a MCS device. As the largest registry
for data on patients receiving MCS devices, INTERMACS has been leveraged
to develop numerous risk prediction models for mortality and other types
of adverse events that may occur after receiving a device
\cite{hsich2012should, cotts2014predictors, eckman2011survival, kirklin2017eighth, kormos2019society, Adamo950}.
An important step for developing risk prediction models is dealing with
missing data, which is common in a ``real world'' database that is not
designed for the rigor of data completeness found in a clinical trial.
While some missing values are related to data entry error, others may
relate to instability of a patient's circulatory status at the time of
implant. For instance, hemodynamic laboratory values are missing for
patients in whom invasive catheterizations were not performed. Patients
without invasive catheterizations may be too ill (unstable) to tolerate
the procedure. Instead, such patients may, after determination of basic
hemodynamics, proceed directly to device placement.

@@

Selecting an optimal strategy to impute missing values such as random
forests can impact the prognostic accuracy of downstream risk prediction
models. In the current analysis, conducting multiple imputation using
random forests emerged as an optimal strategy to impute missing values
in the INTERMACS data. This investigation can directly inform future
analyses of INTERMACS data and provide evidence quantifying the benefit
of imputing missing data with sound methodology.
